---
title: "2023 Rugby World Predictor"
author: "Peter Meihuizen"
date: "2023-06-23"
output: html_document
---
Introduction:

For this project I will attempt to predict the winner of the 2023 Rugby World Cup using historical data of every international rugby match played by between the the historically best 10 international men's rugby teams in the world. The fixtures for the 2023 World Cup has already been released, with it consisting of group stages and knockout matches. Using the past match results from previous years, I will predict who will win every world cup game between the top 10 teams in the world, including the Rugby World Cup Final.

In order to conduct my analysis, I need to import the data set from excel format. This data set includes all matches which have been played since 1984, 3 years before the start of the first Rugby World Cup. 

```{r}
library(readxl)
df <- read_excel("C:\\Users\\Peter Meihuizen\\Desktop\\Masters\\1st Semester\\Data Science\\World Cup Project\\Data\\Rugby_matches_data_1984-2023.xlsx")
```

I am going to do is generate 4 new columns, in order to identify the winner of each game, the points difference of each game and the winner of the World Cup. First I create a function to generate the points difference between the home_team and away_team.

```{r}
# Points difference
library(tidyverse)

df <- df %>% 
  mutate(points_diff = subtract(., "home_score", "away_score"))
```

Next is a variable for the winner and loser of the game.

```{r}
#Winner of game 
df <- df %>% 
  mutate(Winner = ifelse(points_diff > 0, home_team, ifelse(points_diff < 0, away_team, ".")),
         Loser = ifelse(points_diff < 0, home_team, ifelse(points_diff > 0, away_team, ".")))
```

Then a variable which indicates the winner of the World Cup

```{r}
# Winner of the World Cup
# Create a new column WC_winner

df <- df %>%
  mutate(WC_winner = NA) %>%
  {
    for (i in 1:(nrow(.) - 1)) {
      current_row <- .[i, ]
      next_row <- .[i + 1, ]
      
      if (current_row$world_cup & !next_row$world_cup) {
        .$WC_winner[i] <- current_row$Winner
      }
    }
    .
  }

```

I can now see the winners of each world cup up to date

```{r}
WC_Winners <- df %>%
  filter(!is.na(WC_winner)) %>%
  select(competition, WC_winner) %>%
  tibble()


# Print the summary
WC_Winners
```

Ok now I want to see what the win-loss percentage was for every world cup winning team in the 4 years leading up to their respective world cup victory. In order to do this I will look at the number of wins over the combined number of wins and losses over each 4 year period preceding the beginning of the world cup. In 

```{r}

```






Next I want to include the World rankings of each team which have been collected since the conclusion of the 2003 Rugby World Cup. I need to web scrape this information.


```{r}
#Scraping the data
library(rvest)
url <- "https://commons.wikimedia.org/wiki/Data:Men%27s_World_Rugby_rankings.tab"
page <- read_html(url)
table_data <- page %>% html_table(fill = TRUE)
```

Now I turn that list which I scraped into a data frame and clean the data frame to only include the teams from the original data set.

```{r}
library(dplyr)
rankings <- bind_rows(table_data)

#delete unnecessary information in first two rows
rankings <- rankings[-c(1:2), ]

#delete irrelevant teams
rankings <- select(rankings, matches("^(date|New_Zealand|Australia|South_Africa|France|England|Ireland|Wales|Scotland|Italy|Argentina)$"))
```

Next I need to change the date variables from characters to dates in order to merge the ranking data frame with the main data frame (df).

```{r}
df$date <- as.Date(df$date, format = "%Y-%m-%d")
rankings$date <- as.Date(rankings$date, format = "%Y-%m-%d")
```

The rankings stand as such as of the 12 of June 2023.

```{r}
library(tibble)

June_12_2023_rankings <- rankings[rankings$date == "2023-06-12", ]

# Remove the "date" column
June_12_2023_rankings <- June_12_2023_rankings[, -1]

# Convert the row into a vector
values <- (unlist(June_12_2023_rankings))

# Sort the values in increasing order
sorted_values <- sort(values, decreasing = FALSE)

# Create a list of values and column names
result_tibble <- tibble(Column_names = column_name$sorted_values, Value = as.character(sorted_values))

# Print the result list
print(result_tibble)

```


Then I need to merged the data sets to put the rankings of each team into the main data frame.

```{r}
June_12_2023_rankings <- rankings[rankings$date == "2023-06-12", ]

# Remove the "date" column
selected_row <- selected_row[, -1]

# Convert the row into a vector
values <- as.numeric(unlist(selected_row))

# Sort the values in descending order
sorted_values <- sort(values, decreasing = TRUE)

# Sort the column names accordingly
sorted_names <- names(sorted_values)

# Plot the values and column names in descending order
barplot(sorted_values, names.arg = sorted_names, horiz = TRUE, las = 1, main = "Rankings on 2023-06-12")

```

